---
title: "100 Bushels of Corn, Revisited"
format: html
---


About the authors

John Mount is a data scientist based in San Francisco, with 20+ years of experience in machine learning, statistics, and analytics. She is the co-founder of the data science consulting firm Win-Vector LLC, and (with Nina Zumel) the co-author of Practical Data Science with R, now in its second edition.

Nina Zumel is a data scientist based in San Francisco, with 20+ years of experience in machine learning, statistics, and analytics. She is the co-founder of the data science consulting firm Win-Vector LLC, and (with John Mount) the co-author of Practical Data Science with R, now in its second edition.


## Introduction

Nina Zumel presented the "100 Bushels of Corn" puzzle [here](https://rworks.dev/posts/100bushels/) as a fun example of using `R` as a calculator. What if we want `R` to solve the puzzle for us, instead of merely being a calculator?

Let's give that a go.


## Setting up the problem

> 100 bushes of corn are distributed to 100 people such that every man receives 3 bushels, every woman 2 bushels, and every child 1/2 a bushel. How many men, women, and children are there?

We can write the 100 Bushels of Corn problem as finding integer vectors `x` that satisfy `a %*% x = b` for the following `a, b`. The first row specifies the constraint on the total number of men, women and children; the second row specifies the constraint on how many bushels each person gets. Notice that we doubled the values of the second equation to keep everything integral.

```{r}
a <- matrix(c(
   1, 1, 1,
   6, 4, 1),
   nrow=2, ncol=3, byrow = TRUE)

a
```

```{r}
b <- as.matrix(c(100, 200))

b
```

There are at least two main ways to solve this:

  * Brute force. This is kind of the point of computers and programming languages.
  * Linear algebra, in particular linear algebra over the ring of integers. This approach shows some of the richness of the `R` package environment curated at [CRAN](https://cran.r-project.org).
  
Let's take a quick look at these two solution styles.

## Brute Force Solution

A brute force solution is as follows. First, we get a simple upper bound on each variable. We can do this by checking one row of `a %*% x = b`.

```{r}
upper_bounds <- floor(b[2] / a[2, ])
upper_bounds
```

Now we try all plausible solutions.

```{r}
for (x1 in 0:upper_bounds[[1]]) {
  for (x2 in 0:upper_bounds[[2]]) {
    # use a row of a to solve for x3
    x3 <- (b[1] - (a[1, 1] * x1 + a[1, 2] * x2)) / a[1, 3]
    # check constraints
    if ((x3 >= 0) && (abs(x3 %% 1) < 1e-8)) {
      x = as.matrix(c(x1, x2, x3))
      if (all(a %*% x == b)) {
        print(c(x1, x2, x3))
      }
    }
  }
}
```

And this gives us exactly the 7 solutions Nina found. This is some of the magic of having a programmable computer: one can cheaply try a lot of potential solutions without needing a lot of theory.


## Ring Theory Solution

It turns out there is a systematic way to find all of the integral solutions to a linear system quickly, at least for low dimensional solution spaces. To do this we will use a ring theory or linear algebra matrix factorization called the [Hermite normal form](https://en.wikipedia.org/wiki/Hermite_normal_form).  Fortunately, `R` has a package for this, called  [`numbers`](https://CRAN.R-project.org/package=numbers). We attach this package as follows.


```{r}
library(numbers)
```


This package will find for us a lower diagonal integer matrix `h` and a square unimodular matrix `u` such that `h = a %*% u`. Unimodular matrices map the space of integer vectors <code>Z<sup>n</sup></code> to the same space of integer vectors in a 1 to 1, onto, and invertible manner. This means finding integer solution vectors to `a %*% x = b` is equivalent to the problem of finding integer solutions to `h %*% y = b`, where `x = u %*% y`. This second problem is easier, as `h` is lower diagonal- so solving this system is just a matter of "back filling".

Let's first find `h, u`.

```{r}
hnf <- hermiteNF(a)
h <- hnf$H
u <- hnf$U
stopifnot(all(h == a %*% u))
```

```{r}
# lower triangular transform of a
h
```

```{r}
# unimodular transform
u
```

We now have `a %*% u = h`. `a %*% x = b` implies `h %*% y = b` where `x = u %*% y`. Let's solve for a specific solution `xs` by back substitution.

```{r}
# back substitute to solve h %*% y = b
# this uses the fact that h is lower-triangular
h_rank <- sum(diag(h) != 0)
stopifnot(h_rank > 1)
y <- numeric(ncol(a))
y[1] <- b[1] / h[1, 1]
for (i in 2:h_rank) {
      y[i] <- (b[i] - sum(h[i, 1:(i-1)] * y[1:(i-1)])) / h[i, i]
}
stopifnot(all(b == h %*% y))
xs <- u %*% y
stopifnot(all(a %*% xs == b))
xs
```

It is a standard result of linear algebra that all solutions of `a %*% x = b` are of the form `x = xs + z` where `a %*% z = 0` (that is, `z` is in the null space of `h` and `a`). In our case, the null space is spanned by the last column of `u`.

Let's show this column.

```{r}
null_basis <- u[, (h_rank + 1):ncol(u), drop=FALSE]

null_basis
```

So in our case: all integer solutions of `a %*% x = b` are of the form `[500, -800, 400] + k * [-3, 5, -2]` for integer `k`. 

Now we just need to pick `k` to make everything non-negative (an implicit puzzle condition!). The sign changes of entries of `[500, -800, 400] + k * [-3, 5, -2]` happen for `k` where one of the coordinates is equal to zero. These are: 

* `500 - 3 * k == 0`
* `-800 + 5 * k == 0`, and 
* `400 - 2 * k == 0`. 

So the `k` of interest are in the range `ceiling(max(0, 800/5)) <= k <= floor(min(500/3, 400/2))`, or `160 <= k <= 166`. This gives us:

```{r}
for (k in 160:166) {
  soln <- xs + k * null_basis
  print(c(soln[1, 1], soln[2, 1], soln[3, 1]))
}
```

And these are again exactly the solutions Nina Zumel gave [here](https://rworks.dev/posts/100bushels/).

## Conclusion

`R` gives the ability to exploit any combination of innate ability and knowledge, borrowed ability, or brute force in solving problems.

