---
title: "Stage II OSCC — Health Economics Model"
subtitle: "Analysis from the patient's perspective"
author: "Joseph Rickert"
date: 20256-03-15
description: "This post presents a proof of concept for Healthcare Economics Models from the patient's perspective. It constructs an eight-state continuous time Markov Chain from synthetic data based on jump chain probabilities abstracted from the literature. The Markov chain models the health states that patients who are treated either by surgery or definitive radiation therapy will likely visit while being treated for Stage II Oral Squamous Cell Carcinoma. The total times that patients are expected to spend in each health state for each of the two treatment arms are married to EQ-5D utility estimates to produce Quality Adjusted Life Years (QALYs) that can be used to compare how a synthetic cohort of patients would value the two treatments. It is envisioned that this kind of model would be useful for physicians and patient care teams both in formalizing and documenting the treatment decision process and in communicating with patients. The model was constructed with the help of the Posit AI assistant for the RStudio IDE."
image: ""
image-alt: ""
categories: ""
editor: source
format: html
---

Most health care economics models are constructed from the perspective of a managed health care system such as those offered in Canada and several European countries, or from the perspective of some other third party such as an insurance company. Although the benefits of constructing models from the patient's perspective have been discussed in the literature, see for example [Ioannidis & Garber (2011)](https://pmc.ncbi.nlm.nih.gov/articles/PMC3134464/) and [Tai et al. (2016)](https://pubmed.ncbi.nlm.nih.gov/27712720/), few models have been published. Obvious reasons for this situation would include the privacy issues associated with obtaining patient specific data and the apparent lack of economic incentives to commit resources required to create individualized analytical models.

In this post, I explore the position that detailed, patient-specific models may not be necessary in order to achieve most of the benefits of a health economics model from the patients perspective and present a lightweight, cohort model that I believe captures the majority of the benefits to both physicians and patients that might result from a patient focused model.

The model is intended to be a proof of concept, a minimal viable model that may be useful to physicians both in arriving at their decision of how to treat a patient with stage II oral squamous cell carcinoma (OSCC) and in communicating the consequences of their decision to patients.

**Please note that this document and the model it contains is not being presented as a medical analysis. No medical experts have been consulted in its construction. It is merely being offered as an example of what could be done to build a health care economics model from a patient's perspective with today's modeling tools. Also note that I made significant use of [Posit Assistant](https://posit.co/blog/introducing-ai-in-rstudio/) for the RStudio IDE configured with Claude Sonnet 4.6 which I found to be immensely helpful both for code construction and literature searches. However early, unpublished drafts of this post suffered from AI inputs associated with apparently hallucinated documents**

## Overview of the Model

The model begins with the decision to select radiation or surgery and identifies eight health states that patient will pass through. The states visited and the order of visiting them depend on the initial decision. The next step uses the theory of continuous time Markov chains to construct a synthetic data set for a cohort of patients who vary in age and tumor size. The parameters to build the data, estimates for the jump-chain transition probabilities and mean state occupancy times are estimated from the medical literature. It is envisioned will tune these estimates based on the circumstances of individual patients.

Empirical survival curves and the time spent in each state are then calculated from the synthetic data set. Survival curves from the literature are used to calibrate the model. The value of the health state from the patient's point of view are estimated from EQ-5D values published in the literature.

## Continuous Time Markov Chain Simulation

This section builds a synthetic patient cohort for Stage II (T2 N0 M0) oral
squamous cell carcinoma using a continuous-time Markov chain (CTMC). The
simulation is driven by two clinician-interpret able inputs:

- **Jump chain probabilities** `jump_P[i,j]`: given that a patient *leaves*
  state *i*, the probability they transition directly to state *j*.
- **Mean sojourn times** `mean_sojourn[i]`: the average time (months) a patient
  spends in state *i* before transitioning.

The sources for these quantities are identified by comments in the code. The complete references for these sources are given in the Reference section below.


```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Packages used throughout the post"
library(ggplot2)
library(grid)
library(dplyr)
library(msm)
library(gt)
```


### State Diagram

It is my understanding that although most patients wit stage II OSCC are treated with surgery there are times when surgery is either not possible or not preferable and the in these exceptional cases, radiation treatment is the primary alternative. The following state diagram is my attempt to abstract both the health states a patient will experience following treatment and the probable paths that will be taken among them. Eight states seemed to me to be the minimum number of states required to represent the complexities of treatment.


```{r}
#| message: FALSE
#| warning: FALSE
#| code-fold: TRUE
#| code-summary: "Show the code"
# library(ggplot2)
# library(grid)

BG <- "#F0F3F7"

# ── 1. Nodes ─────────────────────────────────────────────────────────────────
# Layout: left=initial treatment, center=adjuvant/NED, right=outcomes/death
nodes <- data.frame(
  id    = 1:8,
  label = c(
    "S1\nSurgery\n± Neck Dissection",
    "S2\nDefinitive\nRadiation",
    "S3\nPost-op\nSurveillance",
    "S4\nAdjuvant RT\n(PORT)",
    "S5\nAdjuvant\nChemoRT (POCRT)",
    "S6\nNED\n(Health)",
    "S7\nLocoregional\nRecurrence",
    "S8\nDeath"
  ),
  x    = c(1.5,  1.5,  4.5,  4.5,  4.5,  7.0,  9.5,  11.5),
  y    = c(8.5,  3.0,  9.5,  6.5,  3.5,  6.5,  7.5,   5.5),
  fill = c("#E67E22","#8E44AD","#F39C12","#2980B9","#16A085",
           "#27AE60","#C0392B","#2C3E50"),
  stringsAsFactors = FALSE
)

# ── 2. Edge definitions: from, to, label, curvature, nudge_x, nudge_y ────────
edge_defs <- list(
  # Surgery → downstream
  c(1, 3,  "No high-risk\nfeatures",    0.00,  0.10,  0.30),
  c(1, 4,  "High-risk:\nmargins/PNI/\nLVI/nodes",  0.10, -0.20,  0.20),
  c(1, 5,  "ECE/pos\nmargins",          0.18,  0.00, -0.30),
  c(1, 8,  "Peri-op\ndeath",           -0.22,  0.10, -0.20),
  # Definitive RT → downstream
  c(2, 6,  "NED",                       0.18,  0.00,  0.25),
  c(2, 7,  "Treatment\nfailure",        0.00,  0.00,  0.25),
  c(2, 8,  "Death",                     0.14, -0.10, -0.20),
  # Post-op Surveillance → downstream
  c(3, 6,  "",                          0.00,  0.10,  0.25),
  c(3, 7,  "",                          0.22,  0.20, -0.10),
  c(3, 8,  "",                         -0.18,  0.10, -0.18),
  # PORT → downstream
  c(4, 6,  "",                          0.00,  0.10,  0.25),
  c(4, 7,  "",                          0.12,  0.10,  0.18),
  c(4, 8,  "",                          0.14,  0.10, -0.18),
  # POCRT → downstream
  c(5, 6,  "",                          0.00,  0.10,  0.25),
  c(5, 7,  "",                          0.00,  0.10,  0.22),
  c(5, 8,  "",                          0.12,  0.10, -0.18),
  # NED → outcomes
  c(6, 7,  "LR recurrence",             0.00, -0.10,  0.25),
  c(6, 8,  "Death",                     0.18,  0.10, -0.14),
  # Locoregional Recurrence → outcomes
  c(7, 8,  "Death",                     0.00,  0.00,  0.25),
  c(7, 6,  "Salvage\nsuccess",         -0.28, -0.28,  0.00)
)

edges <- do.call(rbind, lapply(edge_defs, function(r) {
  data.frame(from=as.integer(r[[1]]), to=as.integer(r[[2]]),
             prob=r[[3]], curvature=as.numeric(r[[4]]),
             nudge_x=as.numeric(r[[5]]), nudge_y=as.numeric(r[[6]]),
             stringsAsFactors=FALSE)
}))

# Attach node coordinates
edges <- merge(edges, nodes[, c("id","x","y")], by.x="from", by.y="id", sort=FALSE)
names(edges)[names(edges)=="x"] <- "x0"; names(edges)[names(edges)=="y"] <- "y0"
edges <- merge(edges, nodes[, c("id","x","y","fill")], by.x="to", by.y="id", sort=FALSE)
names(edges)[names(edges)=="x"] <- "x1"; names(edges)[names(edges)=="y"] <- "y1"
names(edges)[names(edges)=="fill"] <- "dest_col"

# Shorten endpoints to node perimeter
NODE_R <- 0.58
shorten <- function(x0,y0,x1,y1,r) {
  dx<-x1-x0; dy<-y1-y0; d<-sqrt(dx^2+dy^2)
  list(xs=x0+dx*r/d, ys=y0+dy*r/d, xe=x1-dx*r/d, ye=y1-dy*r/d)
}
segs     <- Map(shorten, edges$x0, edges$y0, edges$x1, edges$y1, MoreArgs=list(r=NODE_R))
edges$xs <- sapply(segs,`[[`,"xs"); edges$ys <- sapply(segs,`[[`,"ys")
edges$xe <- sapply(segs,`[[`,"xe"); edges$ye <- sapply(segs,`[[`,"ye")
edges$lx <- (edges$xs+edges$xe)/2 + edges$nudge_x
edges$ly <- (edges$ys+edges$ye)/2 + edges$nudge_y

# ── 3. Build layers ───────────────────────────────────────────────────────────
arrow_layers <- lapply(seq_len(nrow(edges)), function(i) {
  e <- edges[i,,drop=FALSE]
  geom_curve(data=e, mapping=aes(x=xs,y=ys,xend=xe,yend=ye),
             curvature=e$curvature[[1]], colour=e$dest_col[[1]],
             linewidth=1.0, alpha=0.80,
             arrow=arrow(length=unit(9,"pt"), type="closed", ends="last"),
             show.legend=FALSE)
})

label_layers <- lapply(seq_len(nrow(edges)), function(i) {
  e <- edges[i,,drop=FALSE]
  if (nchar(trimws(e$prob[[1]])) == 0) return(NULL)
  geom_label(data=e, mapping=aes(x=lx,y=ly,label=prob),
             colour=e$dest_col[[1]], fill=BG, size=2.6, fontface="bold",
             label.size=0.22, label.r=unit(0.10,"lines"),
             label.padding=unit(0.15,"lines"), show.legend=FALSE)
})
label_layers <- Filter(Negate(is.null), label_layers)

# ── 4. Base plot ──────────────────────────────────────────────────────────────
base_plot <- ggplot() +
  theme_void(base_size=10) +
  theme(
    plot.background  = element_rect(fill=BG, colour=NA),
    panel.background = element_rect(fill=BG, colour=NA),
    plot.title  = element_text(family="serif", face="bold", size=16,
                               colour="#1A2535", hjust=0.5, margin=margin(b=4)),
    plot.subtitle = element_text(size=9, colour="#555555", hjust=0.5,
                                 margin=margin(b=10))
  ) +
  coord_fixed(xlim=c(0, 13), ylim=c(1.5, 11), clip="off") +
  labs(title="Stage II Oral Squamous Cell Carcinoma",
       subtitle="CTMC State Transition Diagram")

# ── 5. Assemble ───────────────────────────────────────────────────────────────
all_layers <- c(arrow_layers, label_layers,
  list(
    # Column labels
    annotate("text", x=1.5,  y=10.8, label="Initial\nTreatment",
             size=3.2, colour="#555", fontface="italic", hjust=0.5),
    annotate("text", x=4.5,  y=10.8, label="Post-surgical\nPathway",
             size=3.2, colour="#555", fontface="italic", hjust=0.5),
    annotate("text", x=7.0,  y=10.8, label="No Evidence\nof Disease",
             size=3.2, colour="#555", fontface="italic", hjust=0.5),
    annotate("text", x=9.5,  y=10.8, label="Disease\nProgression",
             size=3.2, colour="#555", fontface="italic", hjust=0.5),
    annotate("text", x=11.5, y=10.8, label="Absorbing\nState",
             size=3.2, colour="#555", fontface="italic", hjust=0.5),
    # Nodes
    geom_point(data=nodes, aes(x=x, y=y, colour=fill),
               size=28, show.legend=FALSE),
    scale_colour_identity(),
    geom_text(data=nodes, aes(x=x, y=y, label=label),
              colour="white", size=2.8, fontface="bold", lineheight=0.9),
    # Surgical candidate bracket
    annotate("segment", x=0.25, xend=0.25, y=2.2, yend=9.3,
             colour="#888", linewidth=0.5),
    annotate("segment", x=0.25, xend=0.45, y=9.3, yend=9.3,
             colour="#888", linewidth=0.5),
    annotate("segment", x=0.25, xend=0.45, y=2.2, yend=2.2,
             colour="#888", linewidth=0.5),
    annotate("text", x=0.0, y=5.75, label="Surgical\ncandidate?\nYes ↑\nNo ↓",
             size=2.8, colour="#666", hjust=0.5, lineheight=0.9)
  )
)

Reduce("+", all_layers, init=base_plot)
```



| State | Label                   | Role                                              |
|-------|-------------------------|---------------------------------------------------|
| S1    | Surgery                 | Initial surgical treatment (days–weeks)           |
| S2    | DefinitiveRT            | Non-surgical candidates; 6–7 week RT course       |
| S3    | PostOpSurveillance      | Surgery, no high-risk pathological features       |
| S4    | AdjuvantRT (PORT)       | High-risk: margins, PNI, LVI, multiple nodes      |
| S5    | AdjuvantChemoRT (POCRT) | Highest-risk: ECE or positive margins             |
| S6    | NED                     | No evidence of disease; long-term surveillance    |
| S7    | LR Recurrence           | Locoregional recurrence at primary or nodes       |
| S8    | Death                   | Absorbing state                                   |


## Data Simulation

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Simulation setup"
set.seed(6413)

N_PATIENTS <- 400
MAX_TIME   <- 60      # months (5-year follow-up)
obs_base   <- c(0, 1, 2, 3, 6, 9, 12, 18, 24, 36, 48, 60)

state_labels_ii <- c(
  "Surgery", "DefinitiveRT", "PostOpSurveillance",
  "AdjuvantRT_PORT", "AdjuvantChemoRT_POCRT",
  "NED", "LR_Recurrence", "Death"
)

# ── Patient covariates ────────────────────────────────────────────────────────
# Stage II OSCC (T2 N0 M0): tumor 2-4 cm, DOI 5-10 mm
age           <- as.integer(round(rnorm(N_PATIENTS, mean = 62, sd = 11)))
age           <- pmax(32L, pmin(85L, age))
tumor_size_cm <- round(runif(N_PATIENTS, 2.0, 4.0), 1)
DOI_mm        <- round(runif(N_PATIENTS, 5.0, 10.0), 1)

# Surgical candidacy: logistic model, ~78% surgery at age 60
surgery_prob <- plogis(1.5 - 0.04 * (age - 60))
treatment    <- ifelse(runif(N_PATIENTS) < surgery_prob, "Surgery", "DefinitiveRT")
```

### Clinician Inputs

A continuous time Markov chain may be described by its jump chain and transition rate matrix. The jump chain may be thought of as an embedded discrete time Markov chain that gives the probabilities of which state the chain will jump to next when it is time to jump. The transition rates are mean rates at which of the probability distributions that determine when jumps will be made. The major assumption in this model is that the jump probabilities and the times that patients are likely to spend in each state which are inverted to obtain the transition rates will be intuitive to most practicing physicians. The values for these parameters have been derived from the literature, however the model has been set up so that they can be easily changed if the physician disagrees with the with the values in general or thinks that a particular class of patients would be better represented by other values.

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Clinician inputs: jump probabilities and mean sojourn times"

# == CLINICIAN INPUTS ==========================================================
#
# jump_P[i, j]  : probability of moving to state j *given* a transition occurs
#                 from state i.  Rows must sum to 1 for states 1-7.
#
# mean_sojourn  : average time (months) in each transient state.
#
# Together these fully specify the CTMC generator:
#   Q[i, j] = jump_P[i, j] / mean_sojourn[i]   (off-diagonal)
# All downstream simulation code derives Q_base automatically from these inputs.

# ── Jump chain transition probabilities ───────────────────────────────────────


jump_P <- matrix(0, 8, 8,
  dimnames = list(paste0("S", 1:8), paste0("S", 1:8)))

# S1 (Surgery) -> {S3=no high-risk, S4=PORT, S5=POCRT, S8=peri-op death}
# ~20-30% high-risk features -> PORT (Tassone et al. 2023, NCDB n=53,503; DOI:10.1002/ohn.205)
# ~10-15% ECE/pos margins -> POCRT; peri-operative mortality ~1-2% (Nathan et al. 2025, NSQIP n=866; DOI:10.18203/issn.2454-5929.ijohns20252980)
jump_P[1, c(3, 4, 5, 8)] <- c(0.585, 0.25, 0.15, 0.015)

# S2 (DefinitiveRT) -> {S6=NED, S7=treatment failure, S8=death}
# Definitive RT locoregional control ~60-70% at 2 yrs (Szturz 2017)
jump_P[2, c(6, 7, 8)] <- c(0.60, 0.30, 0.10)

# S3 (PostOpSurveillance) -> {S6=NED, S7=LR recurrence, S8=death}
# 5-yr LR recurrence ~15-20% after R0 surgery alone (Luryi 2014, SEER)
jump_P[3, c(6, 7, 8)] <- c(0.71, 0.21, 0.08)

# S4 (PORT) -> {S6=NED, S7=LR recurrence, S8=death}
# After PORT, 3-yr OS ~71% (95% CI 67-75%) in adjuvant PORT cohort (Hosni et al. 2019, PMID: 30244160)
jump_P[4, c(6, 7, 8)] <- c(0.67, 0.23, 0.10)

# S5 (POCRT) -> {S6=NED, S7=LR recurrence, S8=death}
# After POCRT, locoregional control ~55-60% (Bernier 2004, Cooper 2004)
jump_P[5, c(6, 7, 8)] <- c(0.56, 0.30, 0.14)

# S6 (NED) -> {S7=LR recurrence, S8=death}
# ~75% leave NED via recurrence vs other-cause death (SEER Stage II split)
jump_P[6, c(7, 8)] <- c(0.75, 0.25)

# S7 (LR Recurrence) -> {S6=salvage success, S8=death}
# Salvage success ~15-25% (Goodwin 2000, Temam 2005)
jump_P[7, c(6, 8)] <- c(0.20, 0.80)

# ── Mean sojourn times (months) ───────────────────────────────────────────────
mean_sojourn <- c(
  1.0,   # S1  Surgery: days to pathology result
  2.0,   # S2  DefinitiveRT: standard 6-7 week RT course
  14.0,  # S3  PostOpSurveillance: active surveillance before event
  3.0,   # S4  PORT: ~6-week adjuvant RT + recovery
  4.0,   # S5  POCRT: ~6-week concurrent chemoRT + recovery
  50.0,  # S6  NED: years of disease-free follow-up
  12.5,  # S7  LR Recurrence: median OS after LR recurrence ~12-18 mo
  NA     # S8  Death (absorbing)
)
names(mean_sojourn) <- paste0("S", 1:8)

# == END CLINICIAN INPUTS ======================================================

# ── Derive Q_base from clinician inputs ───────────────────────────────────────
# Q[i,j] = jump_P[i,j] / mean_sojourn[i]
Q_base <- matrix(0, 8, 8)
for (i in 1:7) {
  for (j in which(jump_P[i, ] > 0)) {
    Q_base[i, j] <- jump_P[i, j] / mean_sojourn[i]
  }
}
diag(Q_base) <- -rowSums(Q_base)
```

:::: {.columns}

::: {.column width="45%"}



##### Jump Chain Transition Probabilities

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Show code"
as.data.frame(jump_P)
```

:::

::: {.column width="10%"}

:::

::: {.column width="45%"}



##### Mean Sojourn Times and Implied Rates

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Show code"
data.frame(
  #State        = paste0("S", 1:7),
  Label        = state_labels_ii[1:7],
  Mean_sojourn = mean_sojourn[1:7],
  Total_rate   = round(-diag(Q_base)[1:7], 4)
)
```

:::

::::


### Synthetic Data

Here are the first few rows of the synthetic data set. Note that two covariates: age of patient and tumor size are included.

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "show code"

# ── Patient-specific Q (covariates scale recurrence & mortality rates) ─────────
make_Q <- function(age_i, size_i) {
  Q <- Q_base
  age_eff  <- exp(0.025 * (age_i  - 60))   # older  -> higher mortality
  size_eff <- exp(0.180 * (size_i -  3.0))  # larger -> higher recurrence
  Q[3, 7] <- Q_base[3, 7] * size_eff        # PostOpSurv -> LR Recurrence
  Q[6, 7] <- Q_base[6, 7] * size_eff * age_eff
  Q[6, 8] <- Q_base[6, 8] * age_eff
  Q[7, 8] <- Q_base[7, 8] * age_eff
  diag(Q) <- 0
  diag(Q) <- -rowSums(Q)
  Q
}

# ── Simulate one patient's full CTMC path ─────────────────────────────────────
sim_ctmc <- function(init_s, Q_i, max_t) {
  t_vec <- 0; s_vec <- init_s; s <- init_s; t <- 0
  while (t < max_t && s != 8L) {
    total_rate <- -Q_i[s, s]
    if (total_rate == 0) break
    t <- t + rexp(1L, total_rate)
    if (t > max_t) break
    probs    <- pmax(Q_i[s, ], 0)
    probs[s] <- 0
    s        <- sample.int(8L, 1L, prob = probs)
    t_vec    <- c(t_vec, t); s_vec <- c(s_vec, s)
  }
  list(times = t_vec, states = s_vec)
}

# State at a given observation time (last known state)
state_at <- function(traj, obs_t) {
  traj$states[max(which(traj$times <= obs_t))]
}

# ── Generate panel data ────────────────────────────────────────────────────────
records <- vector("list", N_PATIENTS)

for (i in seq_len(N_PATIENTS)) {
  init_s  <- if (treatment[i] == "Surgery") 1L else 2L
  Q_i     <- make_Q(age[i], tumor_size_cm[i])
  traj    <- sim_ctmc(init_s, Q_i, MAX_TIME)

  jitter  <- c(0, runif(length(obs_base) - 1L, -0.5, 0.5))
  obs_t   <- sort(unique(pmax(0, obs_base + jitter)))

  death_t <- if (8L %in% traj$states)
    traj$times[which(traj$states == 8L)[1L]]
  else Inf

  obs_t <- obs_t[obs_t < death_t]
  if (is.finite(death_t) && death_t <= MAX_TIME)
    obs_t <- c(obs_t, death_t)

  obs_s <- sapply(obs_t, state_at, traj = traj)

  records[[i]] <- data.frame(
    id            = i,
    time          = round(obs_t, 2),
    state         = obs_s,
    age           = age[i],
    tumor_size_cm = tumor_size_cm[i],
    DOI_mm        = DOI_mm[i],
    treatment     = treatment[i],
    stringsAsFactors = FALSE
  )
}

oscc2_data <- do.call(rbind, records)
rownames(oscc2_data) <- NULL
oscc2_data$state_label <- state_labels_ii[oscc2_data$state]

# Save synthetic data
write.csv(oscc2_data, "OSCC_HE_Synthetic_Data.csv", row.names = FALSE)

# First 10 rows
head(oscc2_data, 10)
```

---

### Summary of Data Set

This table shows some of the meta-data describing the synthetic data set.

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Cohort summary"
# Collect components for footnotes
trt_counts <- table(treatment)
died <- oscc2_data |>
  dplyr::filter(state == 8) |>
  dplyr::distinct(id) |>
  nrow()

# Main table: state visit counts
as.data.frame(table(State = oscc2_data$state_label)) |>
  gt() |>
  cols_label(State = "State", Freq = "Observations") |>
  tab_source_note(md(paste0(
    "**Dataset:** ", nrow(oscc2_data), " observations; ",
    N_PATIENTS, " patients"
  ))) |>
  tab_source_note(md(paste0(
    "**Treatment:** Surgery = ", trt_counts["Surgery"],
    "; DefinitiveRT = ", trt_counts["DefinitiveRT"]
  ))) |>
  tab_source_note(md(paste0(
    "**Mortality:** ", died, " / ", N_PATIENTS,
    " patients reached Death (",
    round(100 * died / N_PATIENTS, 1), "%)"
  )))
```
### First Patient Trajectory

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Show trajectory"
oscc2_data[oscc2_data$id == 1,
           c("id", "time", "state", "state_label", "treatment",
             "age", "tumor_size_cm", "DOI_mm")]
```


---

## Survival Analysis

The first plot shows overall survival (Kaplan-Meier) curves for both the pooled data including patients in both treatment arms and the overall survival curves for the individual survival curves. Note that the individual curves for both treatments are better than the combined curve. See [Rich et al. (2014)](https://pmc.ncbi.nlm.nih.gov/articles/PMC3932959/#:~:text=DIFFERENT%20KAPLAN%2DMEIER%20CURVES%20USED%20IN%20CANCER%20LITERATURE&text=The%20main%20difference%20between%20the,in%20disease%2Dspecific%20survival%20curves.) for why this might happen with real data. 

The second plot panel shows the overall survival curves for the each treatment for each of the four possible values of the covariates. A technical note at the end of this post provides results of formal logrank tests to look for significant differences.

I am only showing overall survival curves here because computing the sojourn times of each state for each treatment is what drives the calculation of utilities below in the health economics part of the model. A thorough exploration of survival would also require computing progression free survival curves. (See the reference immediately above.)



```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Kaplan-Meier survival curves"
library(survival)
library(survminer)

# One row per patient: time to death or end of follow-up
km_input <- oscc2_data |>
  dplyr::group_by(id, treatment) |>
  dplyr::summarise(
    time  = max(time),
    event = as.integer(any(state == 8)),
    .groups = "drop"
  )

# Add a Combined arm row for each patient so all three curves
# come from a single survfit object (required by ggsurvplot)
km_all_input <- dplyr::bind_rows(
  km_input,
  dplyr::mutate(km_input, treatment = "Combined")
) |>
  dplyr::mutate(treatment = factor(treatment,
    levels = c("Surgery", "DefinitiveRT", "Combined")))

km_fit <- survfit(Surv(time, event) ~ treatment, data = km_all_input)

ggsurvplot(
  km_fit,
  data        = km_all_input,
  conf.int    = TRUE,
  risk.table  = TRUE,
  palette     = c("#E67E22", "#8E44AD", "#2C3E50"),
  legend.labs = c("Surgery", "Definitive RT", "Combined"),
  xlab        = "Time (months)",
  ylab        = "Survival probability",
  title       = "Overall Survival — Stage II OSCC",
  ggtheme     = theme_bw()
)
```


### Survival by Age and Tumour Size

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "KM curves stratified by age × tumour size"
library(broom)
library(ggplot2)
library(dplyr)

# ── One row per patient with covariate categories ─────────────────────────────
# Age split at 62 (simulation mean); tumour size split at 3 cm (Stage II midpoint)
km_cov <- oscc2_data |>
  dplyr::group_by(id, treatment) |>
  dplyr::summarise(
    time          = max(time),
    event         = as.integer(any(state == 8)),
    age           = first(age),
    tumor_size_cm = first(tumor_size_cm),
    .groups       = "drop"
  ) |>
  dplyr::mutate(
    age_grp  = factor(ifelse(age  < 62,  "Age < 62",   "Age ≥ 62"),
                      levels = c("Age < 62", "Age ≥ 62")),
    size_grp = factor(ifelse(tumor_size_cm < 3.0, "Tumour < 3 cm", "Tumour ≥ 3 cm"),
                      levels = c("Tumour < 3 cm", "Tumour ≥ 3 cm")),
    covariate_group = factor(
      interaction(age_grp, size_grp, sep = ", "),
      levels = c(
        "Age < 62, Tumour < 3 cm",
        "Age < 62, Tumour ≥ 3 cm",
        "Age ≥ 62, Tumour < 3 cm",
        "Age ≥ 62, Tumour ≥ 3 cm"
      )
    )
  )

pal4  <- c("#2166AC", "#92C5DE", "#D6604D", "#B2182B")
labs4 <- c("Age < 62, Tumour < 3 cm",
           "Age < 62, Tumour ≥ 3 cm",
           "Age ≥ 62, Tumour < 3 cm",
           "Age ≥ 62, Tumour ≥ 3 cm")

# ── Tidy survival estimates per arm ──────────────────────────────────────────
tidy_arm <- function(arm, data) {
  df  <- dplyr::filter(data, treatment == arm)
  fit <- survfit(Surv(time, event) ~ covariate_group, data = df)
  broom::tidy(fit, conf.int = FALSE) |>
    dplyr::mutate(
      treatment = arm,
      covariate_group = factor(
        sub("covariate_group=", "", strata),
        levels = levels(data$covariate_group)
      )
    )
}

surv_tidy <- dplyr::bind_rows(
  tidy_arm("Surgery",      km_cov),
  tidy_arm("DefinitiveRT", km_cov)
) |>
  dplyr::mutate(treatment = factor(treatment,
    levels = c("Surgery", "DefinitiveRT"),
    labels = c("Surgery Arm", "Definitive RT Arm")))

# ── Two-panel KM plot ─────────────────────────────────────────────────────────
ggplot(surv_tidy, aes(x = time, y = estimate,
                      colour = covariate_group,
                      group  = covariate_group)) +
  geom_step(linewidth = 0.8) +
  facet_wrap(~treatment, ncol = 2) +
  scale_colour_manual(values = pal4, labels = labs4) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
  scale_x_continuous(limits = c(0, 60), breaks = seq(0, 60, 12)) +
  labs(
    x      = "Time (months)",
    y      = "Survival probability",
    colour = "Age / Tumour Size",
    title  = "Overall Survival by Age and Tumour Size — Stage II OSCC"
  ) +
  theme_bw(base_size = 11) +
  theme(
    legend.position  = "bottom",
    legend.direction = "horizontal",
    legend.text      = element_text(size = 9),
    legend.title     = element_text(size = 10, face = "bold"),
    strip.text       = element_text(size = 11, face = "bold"),
    plot.title       = element_text(size = 12, face = "bold")
  ) +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE))
```
---

### What the curves show:

* In the Surgery arm (`r length(unique(oscc2_data$id[oscc2_data$treatment == 'Surgery']))`), the separation between strata is consistent throughout follow-up: younger patients with smaller tumours fare best (dark blue, ~57% at 5 years), while older patients with larger tumours fare worst (dark red, ~30%). Tumour size appears to drive more separation than age alone.
* In the Definitive RT arm (`r length(unique(oscc2_data$id[oscc2_data$treatment == 'DefinitiveRT']))`), the groups are substantially smaller, which produces the characteristic staircase steps. The ordering is broadly similar, but the curves for Age ≥ 62 with larger tumours (dark red) drop sharply early on.

Note that the covariant effects are encoded in the `make_Q` function used to create the synthetic data as follows:

  * `age_eff  <- exp(0.025 * (age_i  - 60))   # older  -> higher mortality`
  * `size_eff <- exp(0.180 * (size_i -  3.0))  # larger -> higher recurrence`


## Time Patients Spend in Each Health State

The total times that patients are expected to spend in each state over a 60-month horizon drive the economics analysis. The ability to make this calculation is the primary motivation for modeling the treatment regimen as a continuous time Markov chain.

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "State occupancy function"
library(expm)

# Adapted from plot_msm_state_occupancy() (Breast_Cancer_v3.qmd).
# Accepts Q_base directly instead of a fitted msm model, and accepts
# a start_state argument so the two treatment arms can be compared.
plot_state_occupancy <- function(Q,
                                 start_state  = 1,
                                 tmax         = 60,
                                 tstep        = 0.5,
                                 state_labels = NULL,
                                 title        = "State Occupancy Probabilities") {
  n_states    <- nrow(Q)
  start_probs <- rep(0, n_states)
  start_probs[start_state] <- 1
  time <- seq(0, tmax, by = tstep)

  occ <- matrix(NA, nrow = length(time), ncol = n_states)
  colnames(occ) <- paste0("S", seq_len(n_states))
  for (i in seq_along(time))
    occ[i, ] <- start_probs %*% expm(Q * time[i])

  df_long <- as.data.frame(occ) |>
    dplyr::mutate(time = time) |>
    tidyr::pivot_longer(-time, names_to = "state", values_to = "prob")

  if (!is.null(state_labels))
    df_long$state <- factor(df_long$state,
      levels = paste0("S", seq_len(n_states)),
      labels = state_labels)

  ggplot(df_long, aes(x = time, y = prob, colour = state)) +
    geom_line(linewidth = 1.0) +
    labs(x = "Time (months)", y = "Probability",
         title = title, colour = "State") +
    theme_bw(base_size = 12)
}

# Expected time in each state: trapezoid integral of P(t) over 0-60 months
compute_state_time <- function(Q, start_state, tmax = 60, tstep = 0.5) {
  n_states    <- nrow(Q)
  start_probs <- rep(0, n_states)
  start_probs[start_state] <- 1
  time <- seq(0, tmax, by = tstep)
  occ  <- matrix(NA, nrow = length(time), ncol = n_states)
  for (i in seq_along(time))
    occ[i, ] <- start_probs %*% expm(Q * time[i])
  apply(occ, 2, function(p)
    sum(diff(time) * (p[-1] + p[-length(p)]) / 2))
}
```




```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Expected state time by arm"
time_surg <- compute_state_time(Q_base, start_state = 1)
time_def  <- compute_state_time(Q_base, start_state = 2)

data.frame(
  State      = paste0("S", 1:8),
  Label      = state_labels_ii,
  Surgery_mo = round(time_surg, 2),
  DefRT_mo   = round(time_def,  2)
)
```

The state occupancy plots show the probability of being in each state as time progresses. Along with the survival curves these plots should make it possible to calibrate the model to times published in the literature.


:::: {.columns}

::: {.column width="45%"}

#### State Occupancy: Surgery 

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Surgery arm occupancy plot"
plot_state_occupancy(
  Q            = Q_base,
  start_state  = 1,
  state_labels = state_labels_ii,
  title        = "State Occupancy Probabilities — Surgery Arm"
)
```

:::

::: {.column width="10%"}

:::

::: {.column width="45%"}

#### State Occupancy: Definitive RT 

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Definitive RT arm occupancy plot"
plot_state_occupancy(
  Q            = Q_base,
  start_state  = 2,
  state_labels = state_labels_ii,
  title        = "State Occupancy Probabilities — Definitive RT Arm"
)
```

:::

::::

---


## Health Economic Evaluation — Patient Perspective

The health Economics model presented here focus on QALYs calculated with with help of EQ-5D utilities. The utilities ought to be of direct help to patients by providing a measure, however imperfect, of how "on average" others with the same disease have rated the ensemble of unpleasant experiences related to each possible health state. To be truly helpful the story behind each utility needs to be told. The side effects of the treatments experienced in each health state need to be enumerated and presented with the strategies that previous patients have employed to cope with them.

The model does not calculate costs under the assumption that, in most cases, cost would not be directly relevant to patients. Moreover in the United States, such calculations would be impossible. The U.S. health care system is set up so that the cost of treatments are only made known to patients after the treatments have be received.

### Quality-Adjusted Life Years (QALYs)

A Quality-Adjusted Life Year (QALY) combines the quantity and quality of life
into a single metric. One QALY represents one year lived in perfect health;
a year spent in a health state with diminished quality counts for less than
one QALY in proportion to that state's utility weight.

In this model, QALYs are calculated in three steps:

1. **Assign utility weights.** Each of the eight health states is assigned
   an EQ-5D utility value between 0 (death) and 1 (perfect health), drawn
   from peer-reviewed literature on head-and-neck cancer patients.

2. **Estimate time in each state.** The continuous-time Markov chain (CTMC)
   simulation provides the expected number of months the cohort spends in
   each state over a 60-month horizon, separately for the Surgery arm and
   the Definitive Radiation Therapy arm.

3. **Compute utility-weighted life years.** For each state, the time (in months)
   is multiplied by its utility weight and the products are summed across all
   states. Dividing by 12 converts the result to years:

$$
\text{Total QALYs} = \frac{1}{12} \sum_{i=1}^{8} \text{time}_i \times u_i
$$

where $\text{time}_i$ is the expected months in state $i$ and $u_i$ is its
EQ-5D utility. The two treatment arms are evaluated on the same 60-month
horizon, allowing a direct QALY comparison from the patient's perspective.


Health state utility values (HSUVs) map each model state to a preference-based
quality-of-life weight on the 0–1 scale (0 = death, 1 = perfect health).
The weights below are EQ-5D-based defaults drawn from the published literature.
S6 (NED) carries the highest utility among living states and serves as the
reference; a NED-normalized column (NED = 1.00) is included for sensitivity work.



### Utility Values

The sources for the utility values are provided as comments in the code below. Complete references are provided in the Reference section at the end.
```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Define and display utility weights"
# Health state utility values (EQ-5D scale, 0 = death, 1 = perfect health).
# S6 (NED) is the reference (highest) living state.
# Sources: 
# S1 de Almeida et al. 2014; Noel et al. 2015
# S2 Truong et al. / RTOG 0522 2017
# S3 Govers et al. 2016
# S4 Govers et al. 2016; Ramaekers et al. 2011
# S5 Truong et al. / RTOG 0522 2017
# S6 Noel et al. 2015 (reference state)
# S7 Meregaglia & Cairns 2017 (sparse data
# S8 Convention (absorbing state)
utility <- c(
  S1 = 0.60,   # Surgery: acute peri-operative period
  S2 = 0.65,   # DefinitiveRT: during 6-7 week RT course
  S3 = 0.72,   # PostOpSurveillance: recovery phase
  S4 = 0.65,   # PORT: adjuvant RT toxicity
  S5 = 0.60,   # POCRT: concurrent chemoRT, highest acute toxicity
  S6 = 0.82,   # NED: reference state (Noel 2015)
  S7 = 0.55,   # LR Recurrence: limited published data
  S8 = 0.00    # Death: absorbing state
)

df_utility <- data.frame(
  State        = paste0("S", 1:8),
  Label        = state_labels_ii,
  Utility_EQ5D = utility,
  Utility_NED1 = round(utility / utility["S6"], 3),
  QALY_Surgery = round(time_surg * utility / 12, 3),
  QALY_DefRT   = round(time_def  * utility / 12, 3),
  QALY_Diff    = round((time_surg - time_def) * utility / 12, 3)
)
rownames(df_utility) <- NULL

df_utility |>
  gt() |>
  cols_label(
    State        = "State",
    Label        = "Health State",
    Utility_EQ5D = "EQ-5D Utility",
    Utility_NED1 = "NED-Normalised",
    QALY_Surgery = "QALYs — Surgery",
    QALY_DefRT   = "QALYs — Definitive RT",
    QALY_Diff    = "Difference (Surg − RT)"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = State == "S6")
  ) |>
  tab_style(
    style = cell_text(color = "#27AE60", weight = "bold"),
    locations = cells_body(columns = QALY_Diff, rows = QALY_Diff > 0)
  ) |>
  tab_style(
    style = cell_text(color = "#C0392B", weight = "bold"),
    locations = cells_body(columns = QALY_Diff, rows = QALY_Diff < 0)
  ) |>
  tab_spanner(
    label   = "Utility",
    columns = c(Utility_EQ5D, Utility_NED1)
  ) |>
  tab_spanner(
    label   = "QALYs by Treatment Arm",
    columns = c(QALY_Surgery, QALY_DefRT, QALY_Diff)
  ) |>
  grand_summary_rows(
    columns  = c(QALY_Surgery, QALY_DefRT, QALY_Diff),
    fns      = list(Total ~ sum(.)),
    fmt      = ~ fmt_number(., decimals = 3)
  ) |>
  tab_source_note("NED-Normalised: utility relative to S6 NED = 1.00. QALYs = utility-weighted months / 12 over a 60-month horizon. Difference = Surgery − Definitive RT; green = Surgery advantage, red = RT advantage.")
```

```{r}
#| include: false
# ── QALY summary scalars for inline narrative references ─────────────────────
qaly_surg_total <- round(sum(time_surg * utility / 12), 2)
qaly_def_total  <- round(sum(time_def  * utility / 12), 2)
qaly_diff_total <- round(abs(qaly_surg_total - qaly_def_total), 2)
qaly_lr_surg    <- round(time_surg[7] * utility["S7"] / 12, 2)
qaly_lr_def     <- round(time_def[7]  * utility["S7"] / 12, 2)
```


```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Utility bar chart"
#| include: false
ggplot(
  data.frame(
    state   = factor(paste0("S", 1:8), levels = paste0("S", 8:1)),
    utility = utility
  ),
  aes(y = state, x = utility)
) +
  geom_col(fill = "#2980B9", alpha = 0.8) +
  geom_vline(xintercept = utility["S6"], linetype = "dashed",
             colour = "#27AE60", linewidth = 0.8) +
  annotate("text", x = utility["S6"] + 0.01, y = 8.4,
           label = "NED (reference)", hjust = 0,
           colour = "#27AE60", size = 3.5) +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_y_discrete(labels = setNames(state_labels_ii, paste0("S", 1:8))) +
  labs(
    x       = "EQ-5D utility (0 = death, 1 = perfect health)",
    y       = NULL,
    title   = "Health State Utility Values — Stage II OSCC",
    caption = "Dashed line marks NED, the highest-utility living state"
  ) +
  theme_bw(base_size = 12)
```

### QALY Comparison by Treatment Arm

With utility weights and state occupancy times established, QALYs can be
compared across the two treatment arms. The stacked bar chart below shows the
utility-weighted life years accrued in each health state over the 60-month
horizon. Each bar is broken down by health state, so the chart reveals not just
the total QALY difference between arms but *where* that difference arises —
which states drive the advantage and which contribute similarly regardless of
treatment choice. Grand totals are printed above each bar for quick comparison.

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "QALY comparison chart"

# Expected utility-weighted months per state per arm,
# using state occupancy integrals and EQ-5D utility weights.
# Total QALYs = sum(time_in_state_months * utility) / 12
qaly_surg <- time_surg * utility
qaly_def  <- time_def  * utility

state_palette <- c(
  S1 = "#E67E22", S2 = "#8E44AD", S3 = "#F39C12",
  S4 = "#2980B9", S5 = "#16A085", S6 = "#27AE60", S7 = "#C0392B"
)

df_qaly <- data.frame(
  State   = rep(paste0("S", 1:8), 2),
  Label   = rep(state_labels_ii, 2),
  Arm     = rep(c("Surgery", "Definitive RT"), each = 8),
  QALY_mo = c(qaly_surg, qaly_def)
) |>
  dplyr::filter(State != "S8") |>
  dplyr::mutate(
    State = factor(State, levels = paste0("S", 1:7)),
    Arm   = factor(Arm,   levels = c("Surgery", "Definitive RT"))
  )

totals_label <- data.frame(
  Arm   = factor(c("Surgery", "Definitive RT"),
                 levels = c("Surgery", "Definitive RT")),
  total = c(sum(qaly_surg), sum(qaly_def)) / 12,
  label = paste0(round(c(sum(qaly_surg), sum(qaly_def)) / 12, 2), " QALYs")
)

ggplot(df_qaly, aes(x = Arm, y = QALY_mo / 12, fill = State)) +
  geom_col(width = 0.55) +
  geom_text(
    data        = totals_label,
    aes(x = Arm, y = total + 0.06, label = label),
    inherit.aes = FALSE,
    fontface    = "bold", size = 4
  ) +
  scale_fill_manual(
    values = state_palette,
    labels = setNames(state_labels_ii[1:7], paste0("S", 1:7))
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.08)),
    breaks = seq(0, 3, 0.5)
  ) +
  labs(
    x        = NULL,
    y        = "Quality-Adjusted Life Years (QALYs)",
    fill     = "Health State",
    title    = "Expected QALYs by Health State and Treatment Arm",
    subtitle = "60-month horizon; utility weights from EQ-5D literature defaults",
    caption  = "State S8 (Death, utility = 0) excluded; contributes 0 QALYs"
  ) +
  theme_bw(base_size = 12)
```


#### A few things worth noting from the chart:

* Surgery accumulates `r qaly_surg_total` QALYs vs. `r qaly_def_total` QALYs for Definitive RT over the 60-month horizon — a difference of ~`r qaly_diff_total` QALYs.
* The dominant contributor in both arms is NED (green), which is expected given its high utility (`r utility["S6"]`) and long sojourn time.
* Surgery's advantage comes largely from the PostOpSurveillance (orange) segment, which has a better utility (`r utility["S3"]`) than Definitive RT's LR Recurrence (dark red) burden.
* The LR Recurrence segment is visibly larger for the Definitive RT arm (`r qaly_lr_def` QALYs vs. `r qaly_lr_surg`), consistent with the higher recurrence probability built into that arm's jump chain.
* These QALY estimates are model-based (from Q_base × utility) and do not yet incorporate cost.

## Discussion


The intended users of the class of models I am proposing are oncologists and their support teams who are making decisions about patient treatments and explaining the possible consequences of the treatments to their patients. In the first use case, oncologists might find the formality of the model useful in finalizing their decisions end document their decision making process. The formality of having a decision supported by a model should be especially helpful in situations where multiple physicians collaborate on deciding treatment options. At the very least a model would be useful in verifying shared assumptions.

A second use of the model would be to help physicians communicate with patients about the consequences of various treatment options. The opportunity here would be for physicians to translate the EQ-5D QALYs into their expectations of what patients are likely to experience in each health state. 

## Technical Notes

### Log-Rank Tests

A formal analysis of the survival times of the covariate groups should include log rank tests or a Bayesian alternative. I have placed the log rank tests here for those interested in this particular data set.

:::: {.columns}

::: {.column width="45%"}

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Log-rank tests for survival arm"
library(gt)

# ── Helper: tidy survdiff output into a display table ─────────────────────────
tidy_survdiff <- function(sd_obj, arm_label) {
  strata_names <- c(
    "Age < 62, Tumour < 3 cm",
    "Age < 62, Tumour ≥ 3 cm",
    "Age ≥ 62, Tumour < 3 cm",
    "Age ≥ 62, Tumour ≥ 3 cm"
  )
  data.frame(
    Stratum    = strata_names,
    N          = sd_obj$n,
    Observed   = sd_obj$obs,
    Expected   = round(sd_obj$exp, 1),
    `(O-E)^2/V` = round((sd_obj$obs - sd_obj$exp)^2 / diag(sd_obj$var), 3),
    check.names = FALSE
  )
}

# ── Run log-rank tests ────────────────────────────────────────────────────────
diff_surg <- survdiff(Surv(time, event) ~ covariate_group,
                      data = dplyr::filter(km_cov, treatment == "Surgery"))

diff_def  <- survdiff(Surv(time, event) ~ covariate_group,
                      data = dplyr::filter(km_cov, treatment == "DefinitiveRT"))

p_surg_lr <- 1 - pchisq(diff_surg$chisq, df = 3)
p_def_lr  <- 1 - pchisq(diff_def$chisq,  df = 3)

fmt_p <- function(p) ifelse(p < 0.001, "< 0.001", formatC(p, digits = 3, format = "f"))

# ── Surgery arm table ─────────────────────────────────────────────────────────
tidy_survdiff(diff_surg) |>
  gt() |>
  tab_header(
    title    = "Log-Rank Test — Surgery Arm",
    subtitle = paste0(
      "χ² = ", round(diff_surg$chisq, 2),
      " on 3 df,  p = ", fmt_p(p_surg_lr)
    )
  ) |>
  cols_label(`(O-E)^2/V` = html("(O−E)<sup>2</sup>/V")) |>
  fmt_number(columns = c(Expected, `(O-E)^2/V`), decimals = 2) |>
  tab_footnote(
    footnote = "O = observed deaths; E = expected deaths under the null of equal survival.",
    locations = cells_column_labels(columns = Observed)
  ) |>
  tab_footnote(
    footnote = "V = variance of the weighted log-rank statistic.",
    locations = cells_column_labels(columns = `(O-E)^2/V`)
  ) |>
  tab_options(table.width = pct(100))
```

:::

::: {.column width="10%"}

:::

::: {.column width="45%"}


```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Log-rank test forDefinitive RT arm"

# ── Definitive RT arm table ───────────────────────────────────────────────────
tidy_survdiff(diff_def) |>
  gt() |>
  tab_header(
    title    = "Log-Rank Test — Definitive RT Arm",
    subtitle = paste0(
      "χ² = ", round(diff_def$chisq, 2),
      " on 3 df,  p = ", fmt_p(p_def_lr),
      "  (caution: small group sizes)"
    )
  ) |>
  cols_label(`(O-E)^2/V` = html("(O−E)<sup>2</sup>/V")) |>
  fmt_number(columns = c(Expected, `(O-E)^2/V`), decimals = 2) |>
  tab_footnote(
    footnote = "O = observed deaths; E = expected deaths under the null of equal survival.",
    locations = cells_column_labels(columns = Observed)
  ) |>
  tab_footnote(
    footnote = "The smallest stratum has n = 12; the test has limited power in this arm.",
    locations = cells_title(groups = "subtitle")
  ) |>
  tab_footnote(
    footnote = "V = variance of the weighted log-rank statistic.",
    locations = cells_column_labels(columns = `(O-E)^2/V`)
  ) |>
  tab_options(table.width = pct(100))
```

:::

::::

### Some observations on the log-rank test

* The Surgery arm log-rank test is significant (χ² = `r round(diff_surg$chisq, 1)`, df = 3, p = `r fmt_p(p_surg_lr)`), indicating that the four age × tumour-size strata have meaningfully different survival trajectories.
* The dominant contrast is between the youngest/smallest-tumour group (best prognosis) and the oldest/largest-tumour group (worst prognosis), consistent with the covariate effects. 
* The Definitive RT arm test is not significant (χ² = `r round(diff_def$chisq, 1)`, df = 3, p = `r fmt_p(p_def_lr)`). This should be interpreted cautiously: the RT arm contains only `r length(unique(oscc2_data$id[oscc2_data$treatment == 'DefinitiveRT']))` patients split across four strata (smallest n = `r min(diff_def$n)`), giving the test limited power to detect differences even if they exist.


### Using the code with Real Data

The format of the generated synthetic data file is suitable for fitting a continuous time Markov chain using the `R` package `msm`. I have verified that `msm` is able to fit the eight state sample model from the synthetic data. It is reasonable to expect that models of similar complexity could be fit from real patient data.


## References

The references in this section are nether complete nor definitive. The literature is enormous.

### References for Jump chain Transition Probabilities

* [Bernier (204)](https://pubmed.ncbi.nlm.nih.gov/15128894/) "Postoperative irradiation with or without concomitant chemotherapy for locally advanced head and neck cancer", N Engl J Medicine 
* [Cooper et al. (2004)](https://www.nejm.org/doi/full/10.1056/NEJMoa032646#:~:text=We%20are%20indebted%20to%20the%20following%2C%20without,collected%20data;%20and%2C%20in%20particular%2C%20Brian%20A.) "Postoperative Concurrent Radiotherapy and Chemotherapy for High-Risk Squamous-Cell Carcinoma of the Head and Neck", N Engl J Medicine 
* [Goodwin (200)](https://onlinelibrary.wiley.com/doi/10.1097/00005537-200003001-00001) "Salvage surgery for patients with recurrent squamous cell carcinoma of the upper aerodigestive tract: when do the ends justify the means?", Laryngoscope
* [Luryi et al. (2014)](https://journals.sagepub.com/doi/abs/10.1177/0194599814551718) "Positive Surgical Margins in Early Stage Oral Cavity Cancer: An Analysis of 20,602 Cases", Otolaryngology–Head and Neck Surgery
* [Hosni et al. (2019)](https://pubmed.ncbi.nlm.nih.gov/30244160/) "Predictors of Early Recurrence Prior to Planned Postoperative Radiation Therapy for Oral Cavity Squamous Cell Carcinoma and Outcomes Following Salvage Intensified Radiation Therapy", Int J Radiat Oncol Biol Phys. 2019;103(2):363–373. Epub 2018 Sep 21. PMID: 30244160. Cited for: 3-year OS ~71% (95% CI 67–75%) after adjuvant PORT in a cohort of 601 OSCC patients (Princess Margaret Cancer Centre, 2003–2015). This paper supports PORT-arm *survival outcomes* (S4 → S6/S7/S8 transition probabilities).
* [Nathan ey al. (2025)](https://www.ijorl.com/index.php/ijorl/article/view/4724) "The influence of reconstruction following hemiglossectomy on perioperative outcomes", Int J Otorhinolaryngol Head Neck Surg. NSQIP database study (n = 866 hemiglossectomy patients, 2008–2022). Reports perioperative complication rates for OCSCC resection; consistent with ~1–2% 30-day mortality in modern series. Supports revision of jump_P[1,8] from 0.05 to 0.015.
* [Szturz et al. (2017)](https://pubmed.ncbi.nlm.nih.gov/28533474/) "Weekly Low-Dose Versus Three-Weekly High-Dose Cisplatin for Concurrent Chemoradiation in Locoregionally Advanced Non-Nasopharyngeal Head and Neck Cancer: A Systematic Review and Meta-Analysis of Aggregate Data", Oncologist.
* [Tam et al. (2017)](https://jamanetwork.com/journals/jamaotolaryngology/fullarticle/2618943) "Estimating Survival After Salvage Surgery for Recurrent Oral Cavity Cancer". JAMA Otolaryngology Head & Neck Surgery
* [Tassone et al. (2023)](https://aao-hnsfjournals.onlinelibrary.wiley.com/doi/10.1002/ohn.205) "Going Off Guidelines: An NCDB Analysis of Missed Adjuvant Therapy Among Surgically Treated Oral Cavity Cancer", Otolaryngol Head Neck Surg 2023;169(3):632–641. Cited for: clinical framework establishing PORT vs POCRT indications (ECE/positive margins → POCRT; T3–4/N2–3/LVI → PORT) across 53,503 surgically treated oral cavity SCC patients. Supports jump_P[1,4] = 0.25 (PORT allocation) and jump_P[1,5] = 0.15 (POCRT allocation). 
* [Temam et al. (2005)](https://onlinelibrary.wiley.com/doi/abs/10.1002/hed.20234) "Treatment of the N0 neck during salvage surgery after radiotherapy of head and neck squamous cell carcinoma", NECK



### References for EQ-5D Utilities

* [Meregaglia & Cairns (2017)](https://pubmed.ncbi.nlm.nih.gov/28865475/)  "A systematic literature review of health state utility values in head and neck cancer", Health and Quality of life Outcomes. - systematic review of 28 studies and 346 HSUVs in HNC, recommends EQ-5D. 
* [Noel et al. (2015)](https://jamanetwork.com/journals/jamaotolaryngology/fullarticle/2397443#:~:text=Design%2C%20Setting%2C%20and%20Participants%20In,Mark%203%20(HUI3)%20questionnaire.) "Comparison of Health State Utility Measures in Patients With Head and Neck Cancer", JAMA Otolaryngology–Head & Neck Surgery (July 23, 2015), is a prospective, cross-sectional, and longitudinal study involving 100 patients with squamous cell carcinoma of the upper aerodigestive tract. In the prospective cohort,
mean EQ-5D-5L = 0.82 three months post-treatment with no evidence of recurrence.
* [de Almeida et al. (2013)](https://onlinelibrary.wiley.com/doi/abs/10.1002/hed.23340) "Preferences and utilities for health states after treatment for oropharyngeal cancer: Transoral robotic surgery versus definitive (chemo)radiotherapy" Head and Neck, 
* [Govers et al. (2016)](https://onlinelibrary.wiley.com/doi/abs/10.1111/coa.12502) "Quality of life after different procedures for regional control in oral cancer patients: cross-sectional survey", Clinical Otolarngology - EQ-5D-3L in oral cavity OSCC.
* [Truong et al./RTOG 0522 (2017)](https://pmc.ncbi.nlm.nih.gov/articles/PMC5303682/) "Quality of Life and Performance Status From a Substudy Conducted Within a Prospective Phase 3 Randomized Trial of Concurrent Accelerated Radiation Plus Cisplatin With or Without Cetuximab for Locally Advanced Head and Neck Carcinoma: NRG Oncology Radiation Therapy Oncology Group 0522", Int J Radiat Oncol Biol Phys — EQ-5D during concurrent chemoradiotherapy.
